Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-19 prop_letter.tex
prop_letter.tex Reviewing...

In the remainder of this section, we will show that our definition of
QIVPMs is consistent with classical IVPMs in the sense that a
restriction of QIVPMs to commuting events recovers the definition of
classical IVPMs. To formalize this connection between QIVPMs and
classical IVPMs, we first note that in contrast to the case with
conventional quantum probability spaces, there is generally no unique
state~$\rho$ consistent with a QIVPM. We say that a QIVPM $\bar{\mu}$
is consistent with a state $\rho$ and projector~$P$ if the interval
$\bar{\mu}(P)$ contains the exact probability calculated by the Born
rule, i.e, if:
\begin{equation}
\mu_{\rho}^{\mathrm{B}}\left(P\right)\in\bar{\mu}\left(P\right)
\end{equation}
where $\mu_{\rho}^{\mathrm{B}}\left(P\right)=\Tr\left(\rho P\right)$
is the probability that the event~$P$ is determined by the Born
rule~\citep{Born1983,peres1995quantum,544199}.

@@

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the probability
intervals associated with each state~$\rho$ that is consistent with
$\bar{\mu}$ on every projector in the spectral decomposition of the
observable. If $\bar{\mu}$ is a conventional (Born) probability measure
induced by a state $\rho$, then the Born rule probability induced
by every state in $\coreBorn\left(\bar{\mu},\events'\right)$ will
be $\mu_{\rho}^{\mathrm{B}}$ and the interval collapses to a point,
thus reducing the definition to that of Eq.~(\ref{eq:quantum-expectation}).
For example, we can extend one of the fragments of valid probability
measures~$\mu_{1}$ defined in Table~\ref{tab:quantum-probability-measure}
to a QIVPM~$\bar{\mu}_{3}$: $\bar{\mu}_{3}(\proj{0})=\left[\frac{1}{2},\frac{1}{2}\right]$,
$\bar{\mu}_{3}(\proj{1})=\left[\frac{1}{2},\frac{1}{2}\right]$, and
$\bar{\mu}_{3}(\proj{2})=\left[0,0\right]$, ... etc. For the same
observable~$\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$, the only
state~$\rho$ consistent with $\bar{\mu}_{3}$ on $\proj{0}$, $\proj{1}$,
and $\proj{2}$ must be $\proj{\ps}$, and $\mu_{1}=\mu_{\proj{\ps}}^{\mathrm{B}}$.
Therefore, the expectation value~$\expval{\mathbf{O}}_{\bar{\mu}_{3}}=\left[\expval{\mathbf{O}}_{\mu_{1}},\expval{\mathbf{O}}_{\mu_{1}}\right]=\left[1.5,1.5\right]$
which is an interval, but not between $0$ and $1$ because it is
not a probability of an event. We also note that, when restricted
to commuting projectors, the definition above is consistent with the
classical notion of the \emph{Choquet integral}~\citep{GilboaSchmeidler1994,Grabisch2016}
which is used to calculate the expectation value of random variables
as a weighted average.

@@

\noindent The definition puts no restrictions on the choice of intervals,
only on the assignment of intervals to events. When $\delta=0$, every
event must be assigned a probability either in the interval $\imposs$
or in $\necess$, i.e., every event is completely determined with
certainty. As $\delta$ gets larger, the QIVPM allows for more indeterminate
behavior. To illustrate the example easier, we restrict ourselves
to the classical subspace of commuting events~$\events_{0}=\left\{ \mathbb{0},\proj{0},\proj{1},\mathbb{1}\right\} $.
If we apply the same definition in the classical case, the classical
IVPM~$\bar{\mu}_{4}$ defined in Table~\ref{tab:classical-IVPMs}
is $0$-deterministic because $\imposs\subseteq\left[0,0\right]$
and $\necess\subseteq\left[1-0,1\right]$.
\begin{table}
\noindent \centering{}\caption{\label{tab:classical-IVPMs}Classical IVPMs~$\bar{\mu}_{4}$, $\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ on $\events_{0}$.}
\begin{tabular}{cccc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}_{4}(P)$ & $\bar{\mu}_{5}(P)$ & $\bar{\mu}_{6}(P)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\necess$ & $\left[\frac{1}{3},1\right]$ & $\left[\frac{2}{3},\frac{2}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\imposs$ & $\left[0,\frac{1}{3}\right]$ & $\left[\frac{1}{3},\frac{1}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $\necess$ & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By interpreting the states ~$\proj{0}$ and $\proj{1}$ with the
head and the tail of a coin again, $\bar{\mu}_{4}$ represents tossing
a double-headed coin which can be predicted with certain. When $\delta$
is increased to $\frac{1}{3}$, the classical IVPMs~$\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ defined in Table~\ref{tab:classical-IVPMs}
are clearly $\frac{1}{3}$-deterministic, and they have more indeterminate
behavior than $\bar{\mu}_{4}$. However, the most indeterminate classical
IVPM~$\bar{\mu}_{2}$ defined in Eq.~(\ref{eq:classical-no-information-IVPM})
is still not $\frac{1}{3}$-deterministic because $\bar{\mu}_{2}(\proj{0})=\bar{\mu}_{2}(\proj{1})=\unknown$,
but neither $\unknown\subseteq\left[0,\frac{1}{3}\right]$ nor $\unknown\subseteq\left[1-\frac{1}{3},1\right]$.
Finally, when $\delta=1$, because every interval is a subset of $\unknown$,
every QIVPM is $1$-deterministic. In particular, the informativeness
QIVPM~$\bar{\mu}_{1}$ and the Born QIVPM~$\bar{\mu}_{3}$ defined
in Sec.~\ref{sec:Interval-Uncertainty} are both $1$-deterministic.

@@

\begin{thm}[Non-extensible of Gleason's theorem to general QIVPMs]In
a Hilbert space $\Hilb$ of dimension $3$, there is a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$
such that $\coreBorn\left(\bar{\mu},\events\right)=\emptyset$.\end{thm}

Given a three-dimensional Hilbert space $\Hilb$ with an orthonormal
basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $, let $\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} $,
$\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$, $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$,
and \begin{subequations} 
\begin{eqnarray}
 &  & \begin{aligned}\imposs & =\bar{\mu}(\mathbb{0})=\bar{\mu}(\proj{0})\\
 & =\bar{\mu}(\proj{\ps})=\bar{\mu}(\proj{\ps'})\,,
\end{aligned}
\\
 &  & \begin{aligned}\necess & =\bar{\mu}(\mathbb{1})=\bar{\mu}(\mathbb{1}-\proj{0})\\
 & =\bar{\mu}(\mathbb{1}-\proj{\ps})=\bar{\mu}(\mathbb{1}-\proj{\ps'})\,,
\end{aligned}
\\
 &  & \unknown=\bar{\mu}(P)\qquad\textrm{otherwise.}
\end{eqnarray}
\end{subequations} It is straightforward to verify that $\bar{\mu}$
is a QIVPM. Then we will prove $\coreBorn\left(\bar{\mu},\events\right)$
is an empty set by contradiction. Suppose there is a state $\rho=\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}\in\coreBorn\left(\bar{\mu},\events\right)$,
where $\sum_{j=1}^{N}q_{j}=1$ and $q_{j}>0$. Since $\mu_{\rho}^{\mathrm{B}}(P)\in\bar{\mu}(P)$
for any~$P$, the state~$\rho$ must satisfy: 
\begin{equation}
\mu_{\rho}^{\mathrm{B}}(\proj{0})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps'})=0
\end{equation}
which implies $\ip{\phi_{j}}{0}=\ip{\phi_{j}}{\ps}=\ip{\phi_{j}}{\ps'}=0$
for all $j$. However, the last equation never holds. Thus, there
is no state consistent with the QIVPM on all projectors~$P\in\events$,
so Gleason's theorem cannot be extended to general QIVPMs because
of this example.

>>>>

In the remainder of this section, we will show that our definition of
QIVPMs is consistent with classical IVPMs in the sense that a
restriction of QIVPMs to commuting events recovers the definition of
classical IVPMs. To formalize this connection between QIVPMs and
classical IVPMs, we first note that in contrast to the case with
conventional quantum probability spaces, there is generally no unique
state~$\rho$ consistent with a QIVPM~\citep{HOSTunpublished}. We say that a QIVPM $\bar{\mu}$
is consistent with a state $\rho$ and projector~$P$ if the interval
$\bar{\mu}(P)$ contains the exact probability calculated by the Born
rule, i.e, if:
\begin{equation}
\mu_{\rho}^{\mathrm{B}}\left(P\right)\in\bar{\mu}\left(P\right)
\end{equation}
where $\mu_{\rho}^{\mathrm{B}}\left(P\right)=\Tr\left(\rho P\right)$
is the probability that the event~$P$ is determined by the Born
rule~\citep{Born1983,peres1995quantum,544199}.

@@

\noindent Intuitively the expectation value of an observable relative
to a QIVPM $\bar{\mu}$ lies between two possible outcomes, which
themselves lie between the minimum and maximum bounds of the probability
intervals associated with each state~$\rho$ that is consistent with
$\bar{\mu}$ on every projector in the spectral decomposition of the
observable. If $\bar{\mu}$ is a conventional (Born) probability measure
induced by a state $\rho$, then the Born rule probability induced
by every state in $\coreBorn\left(\bar{\mu},\events'\right)$ will
be $\mu_{\rho}^{\mathrm{B}}$ and the interval collapses to a point,
thus reducing the definition to that of Eq.~(\ref{eq:quantum-expectation}).
For example, we can extend one of the fragments of valid probability
measures~$\mu_{1}$ defined in Table~\ref{tab:quantum-probability-measure}
to a QIVPM~$\bar{\mu}_{3}$:
\begin{equation}
\begin{aligned}\bar{\mu}_{3}(\proj{0}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] & \bar{\mu}_{3}(\proj{2}) & =\left[0,0\right]\\
\bar{\mu}_{3}(\proj{1}) & =\left[\tfrac{1}{2},\tfrac{1}{2}\right] &  & \ldots\textrm{ etc.}
\end{aligned}
\label{eq:Born-QIVPM}
\end{equation}
For the same
observable~$\mathbf{O}=\proj{0}+2\,\proj{1}+3\,\proj{2}$, the only
state~$\rho$ consistent with $\bar{\mu}_{3}$ on $\proj{0}$, $\proj{1}$,
and $\proj{2}$ must be $\proj{\ps}$, and $\mu_{1}=\mu_{\proj{\ps}}^{\mathrm{B}}$.
Therefore, the expectation value~$\expval{\mathbf{O}}_{\bar{\mu}_{3}}=\left[\expval{\mathbf{O}}_{\mu_{1}},\expval{\mathbf{O}}_{\mu_{1}}\right]=\left[1.5,1.5\right]$
which is an interval, but not between $0$ and $1$ because it is
not a probability of an event. We also note that, when restricted
to commuting projectors, the definition above is consistent with the
classical notion of the \emph{Choquet integral}~\citep{GilboaSchmeidler1994,Grabisch2016}
which is used to calculate the expectation value of random variables
as a weighted average.

@@

\noindent The definition puts no restrictions on the choice of intervals,
only on the assignment of intervals to events. When $\delta=0$, every
event must be assigned a probability either in the interval $\imposs$
or in $\necess$, i.e., every event is completely determined with
certainty. As $\delta$ gets larger, the QIVPM allows for more indeterminate
behavior. To illustrate the example easier, we restrict ourselves
to the classical subspace of commuting events~$\events_{0}=\left\{ \mathbb{0},\proj{0},\proj{1},\mathbb{1}\right\} $.
If we apply the same definition in the classical case, the classical
IVPM~$\bar{\mu}_{4}$ defined in Table~\ref{tab:classical-IVPMs}
is $0$-deterministic because $\imposs\subseteq\left[0,0\right]$
and $\necess\subseteq\left[1-0,1\right]$.
\begin{table}
\noindent \centering{}\caption{\label{tab:classical-IVPMs}Classical IVPMs~$\bar{\mu}_{4}$, $\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ on $\events_{0}$.}
\begin{tabular}{cccc}
\toprule 
\addlinespace
$P$ & $\bar{\mu}_{4}(P)$ & $\bar{\mu}_{5}(P)$ & $\bar{\mu}_{6}(P)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $\imposs$ & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\necess$ & $\left[\frac{1}{3},1\right]$ & $\left[\frac{2}{3},\frac{2}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\imposs$ & $\left[0,\frac{1}{3}\right]$ & $\left[\frac{1}{3},\frac{1}{3}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $\necess$ & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
By interpreting the states ~$\proj{0}$ and $\proj{1}$ with the
head and the tail of a coin again, $\bar{\mu}_{4}$ represents tossing
a double-headed coin which can be predicted with certain. When $\delta$
is increased to $\frac{1}{3}$, the classical IVPMs~$\bar{\mu}_{5}$
and $\bar{\mu}_{6}$ defined in Table~\ref{tab:classical-IVPMs}
are clearly $\frac{1}{3}$-deterministic, and they have more indeterminate
behavior than $\bar{\mu}_{4}$. However, the most indeterminate classical
IVPM~$\bar{\mu}_{2}$ defined in Eq.~(\ref{eq:classical-no-information-IVPM})
is still not $\frac{1}{3}$-deterministic because $\bar{\mu}_{2}(\proj{0})=\bar{\mu}_{2}(\proj{1})=\unknown$,
but neither $\unknown\subseteq\left[0,\frac{1}{3}\right]$ nor $\unknown\subseteq\left[1-\frac{1}{3},1\right]$.
Finally, when $\delta=1$, because every interval is a subset of $\unknown$,
every QIVPM is $1$-deterministic. In particular, the informativeness
QIVPM~$\bar{\mu}_{1}$ defined in Eq.~(\ref{eq:no-information-IVPM})
and the Born QIVPM~$\bar{\mu}_{3}$ defined in Eq.~(\ref{eq:Born-QIVPM})
are both $1$-deterministic.

@@

\begin{thm}[Non-extensible of Gleason's theorem to general QIVPMs]\label{thm:Non-extensible-of-Gleason's}In
a Hilbert space $\Hilb$ of dimension $3$, there is a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$
such that $\coreBorn\left(\bar{\mu},\events\right)=\emptyset$.\end{thm}

\noindent Given a three-dimensional Hilbert space $\Hilb$ with an
orthonormal basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $, let
$\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} $, $\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$,
$\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$, and \begin{subequations}
\begin{eqnarray}
 &  & \begin{aligned}\imposs & =\bar{\mu}(\mathbb{0})=\bar{\mu}(\proj{0})\\
 & =\bar{\mu}(\proj{\ps})=\bar{\mu}(\proj{\ps'})\,,
\end{aligned}
\\
 &  & \begin{aligned}\necess & =\bar{\mu}(\mathbb{1})=\bar{\mu}(\mathbb{1}-\proj{0})\\
 & =\bar{\mu}(\mathbb{1}-\proj{\ps})=\bar{\mu}(\mathbb{1}-\proj{\ps'})\,,
\end{aligned}
\\
 &  & \unknown=\bar{\mu}(P)\qquad\textrm{otherwise.}
\end{eqnarray}
\end{subequations} It is straightforward to verify that $\bar{\mu}$
is a QIVPM. Then we will prove $\coreBorn\left(\bar{\mu},\events\right)$
is an empty set by contradiction. Suppose there is a state $\rho=\sum_{j=1}^{N}q_{j}\proj{\phi_{j}}\in\coreBorn\left(\bar{\mu},\events\right)$,
where $\sum_{j=1}^{N}q_{j}=1$ and $q_{j}>0$. Since $\mu_{\rho}^{\mathrm{B}}(P)\in\bar{\mu}(P)$
for any~$P$, the state~$\rho$ must satisfy: 
\begin{equation}
\mu_{\rho}^{\mathrm{B}}(\proj{0})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps'})=0
\end{equation}
which implies $\ip{\phi_{j}}{0}=\ip{\phi_{j}}{\ps}=\ip{\phi_{j}}{\ps'}=0$
for all $j$. However, the last equation never holds. Thus, there
is no state consistent with the QIVPM on all projectors~$P\in\events$,
so Gleason's theorem cannot be extended to general QIVPMs because
of this example.

Despite the non-extensible of Gleason's theorem to general QIVPMs,
when we restrict $\bar{\mu}$ on any commuting subspace of events~$\events_{C}$,
$\bar{\mu}$ becomes classical, and Thm.~\ref{thm:Shapley} guarantees
$\coreBorn\left(\bar{\mu},\events_{C}\right)$ is nonempty. In this
sense, Thm.~\ref{thm:Non-extensible-of-Gleason's} is a completely
new phenomena when we generalize both classical IVPMs and quantum
probability measures to QIVPMs.
